---
title: Climate impacts plant fertility in space and time

author: |
  | Halbritter, A. H. ^[Corresponding author: aud.halbritter@uib.no] $^1$, Goldberg, D. $^2$, Gya, R. $^1$, Jaroszynska, F. $^1$, 
  | Klanderud, R. $^3$, Olsen, S. L. $^4$, Telford, R. J. $^1$, Töpper, J. P. $^4$ 
  | and Vandvik, V. $^1$
address: |
  | $^1$University of Bergen, $^2$University of Michigan, $^3$NMBU, $^4$Nina


abstract: |
  Climate change impacts on biodiversity are often inferred from studies investigating species’ performance in response to climatic variability in time or space, i.e. in response to temporal climate fluctuations or along spatial environmental gradients. However, the potential underlying drivers, and hence the interpretation of the observed responses, differs between temporal and spatial gradient studies. Temporal studies are effective in assessing ecological response rates and directions of change, as well as the underlying processes, and, in the case of longer time-series, directions of change and magnitude of time-lags. In contrast, spatial gradient studies reflect long-term outcomes of different climatic regimes, and are therefore suited to assess evolutionary outcomes and ecological equilibria under different climatic conditions. Combining the two approaches in a single study can therefore inform about both short-term responses and longer-term outcomes of climate change on biota. Plant reproduction is generally spatially variable, reflecting long term responses to a specific environmental condition, and at the same time plants respond plastically in reproduction to short term seasonal climatic fluctuations (i.e. if a plant flowers this year or not). Reproductive effort also varies between species, depending on different strategies, such as growth form and clonality, and we hypothesize that graminoids and forbs differ in their responses across space and time. Here, we study plant fertility in 12 grassland sites, selected to represent a spatial ‘climate grid’ with independently varying summer temperature (6.5 - 10.5°C) and annual precipitation (600 - 2200 mm). Throughout this grid, we assessed plant reproductive effort (measured as the frequency of investment in buds, flowers or seeds) in 181 species of vascular plants in 120 permanently marked plots over nine years. The combined spatio-temporal dataset allowed us to assess how the reproductive effort in these plant communities is affected by climate on different time-scales and through different mechanisms. The plant communities differed in fertility across time and space. They were generally less fertile in wet years, and had higher fertility towards warmer and drier climates. Fertility, both in terms of  spatial patterns and plastic responses, differed between graminoids and forbs. Graminoids were less fertile in warm years, while fertility in forbs increased  following a warm year. The combined results of the space and time approaches allowed us to detect climate control where a single approach could have overlooked them, i.e. only looking at temporal changes without accounting for spatial differences.

date: "`r format(Sys.Date(), '%d %B %Y')`"

# These are LaTex settings to take care of floating figures/tables, line spacing, etc
header-includes:
  #- \usepackage{endfloat} put figs at the end
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  
output:
  bookdown::pdf_book:
    toc: no
  #bookdown::pdf_document2:
  fig_caption: yes
  # number_sections: true
  
documentclass: article
classoption: a4paper
bibliography: Paperpile - Jul 07 BibTeX Export.bib
csl: journal-of-ecology.csl

 #always_allow_html: yes

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
options(digits = 2)
#theme_set(theme_bw(base_size = 12))
#knitr::opts_chunk$set(out.width = '100%', dpi=300)

loadd()

```

**Keywords:** fertility, temperature, precipitation, plant, reproduction

## INTRODUCTION
Climate change has cause an alteration in means and variation in both global temperature and precipitation for the last 70 years @Ipcc2013-li. This has led to a change in the abiotic conditions the plants experience with impacts on biodiverstiy and in particular plant distribution, composition, and demography @Elmendorf2012-bz;@Pacifici2015-il;@Topper2018-oy.

Different approaches are used to understand the impacts of climate change on biodiversity and plant performans. We can investigate species’ performance in response to climatic variability in space or time (refs). The potential underlying drivers, and hence the interpretation of the observed responses, differs between temporal and spatial gradient studies. Spatial approaches are studies that exploit the changing abiotic factors along environmental gradients, i.e. elevation, continentality. They reflect long-term outcomes of different climatic regimes, and are therefore suited to assess evolutionary outcomes and ecological equilibria under different climatic conditions. Temporal approaches, assess species responses over time and are effective in assessing ecological response rates and directions of change, as well as the underlying processes, and, in the case of longer time-series, directions of change and magnitude of time-lags (plastic response/no time lag). Combining the two approaches in a single study can therefore inform about both short-term plastic responses and longer-term outcomes of climate change on plant communities.
[//]: # "from https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2435.12599"
However, climatic manipulations need to extend over sufficient spatial scale and numerous years in order to encompass also delayed responses and the complexity and potential feedbacks of interactions (Michalet 2007; Maestre et al . 2009). Such large‐scale and long‐term experiments are costly, time‐consuming and so far very rare, especially in dryland ecosystems (Tielbörger & Salguero‐Gomez 2013).

Plant reproduction varies both spatially across climate gradients (ref) temporally to short-term variability in climatic conditons (i.e. seasonal climatic fluctuations) (ref). Warmer climate can advance and accellerate flowering, while too warm conditions can slow down. Too dry or wet conditions can prevent flowering. In mountains, more precipitation as snow, will shorten the growing season and can again accellerate the flowering for species that are adapted to such conditions. How does the previous year climate affect reproduction?

Other factors than climate can affect reproductive effort in plants. Plant traits/properties such as functional group (forb, graminoid), habitat (alpine/lowland/generalist), clonality, pollination, and longevity can have an effect on reproduction.
- Graminoids are less vulnerable to changes in temperature and precipitation than forbs (ref).
- Alpine plants invest more in energy storage such as seedbanks, clonal growth and the plants only flower in years where the conditons (climate) are benefitial.
- Weather a plant depends on pollinators or is selfing can have an effect on how much a plant needs to flower in one grwoing season.
Finally, life strategy such as longevity (annual/perennial) will define how desperetly a plant needs to flower and reproduce in a year.
Life form and reproduction strategy will affect how plants respond to changes in climate across climate gradients.

Here, we study reproduction in 12 grassland communities, selected to represent a spatial ‘climate grid’ with independently varying summer temperature (6.5 - 10.5°C) and annual precipitation (600 - 2200 mm) over a temporal scale of 10 years. Throughout this grid, we assessed plant reproductive effort (measured as the frequency of investment in buds, flowers or seeds) in `r fertile %>% mutate(species = as.factor(species)) %>% distinct(species) %>% count()` species of vascular plants in `r fertile %>% mutate(species = as.factor(turfID)) %>% distinct(turfID) %>% count()` permanently marked plots over a period of nine years. The combined spatio-temporal dataset allowed us to assess how the reproductive effort in these plant communities is affected by climate on different time-scales and through different mechanisms. Specifically, we ask how does the reproductive effort in these plant communities change in space and time?

Maybe also ask how different life and reproduction strategies affect the reproductive effort in these communities and at the species level (functional group, habitat, clonality, pollination, longevity).


## METHODS
### Study site and plant communities
Twelve semi-natural grassland sites were selected in 2009 in western Norway, within a climate grid consisting of a temperature gradient from boreal to alpine replicated along a precipitation gradient from continental to oceanic (Fig. 1). The sites differ in mean summer temperature (mean temperature of the four warmest month: alpine 6.5°C, subalpine 8.5°C and boreal 10.5°C) and annual precipitation (600, 1200, 2000 and 2700 mm; interpolated climate data with 100m resolution from the normal period 1961 - 1990 provided by the Norwegian Meteorological Institute, met.no). Except for the two climate variables, the sites were selected to be as similar as possible. They are associated with calcareous bedrock, sociological association *Potentillo-Festucetum ovinae* @Fremstad1997-to, south-west facing, and moderately grazed @Klanderud2015-po. The sites were fenced to avoid grazing by larger herbivores and annually mowed to simulate the grazing regime.

```{r Table1}

#loadd(Table1)


```


```{r ClimatePlot, fig.cap = "Gridded climate data in the twelve site over a 11 year period."}

readd(ClimatePlot)


```


### Data collection
In 2009, five blocks with ten randomly placed 25x25cm plots were established. The plots were part of a transplant experiment and half of the plots in each block were transplanted within the block (dug out and placed back in the ground). We use these plots as control plots, because there was never found a difference in species response (community composition, demography,...?) compared to the untouched control plots. Each plot was divided into 25 subplots. In each subplot, fertility (if a plant has produced buds, flower or seeds) was recorded at the peak growing season in 2009, 2011, 2012, 2013, 2015 and 2017. There was no data collection in 2010, 2014 and 2016, because of resting plots, and after 2013, only every second year. We therefore added the same data from control plots from another experiment in the same sites. Here, flowering was recorded only in 6 subplots? Need to check if this makes a difference.
At each site, the local climate data was recorded since 2009. We recorded soil and air temperature...




 



What to do/test:
Look at how snow:rain ration, growing season length affects fertility
Could look at soil temperature, but it correlates with air temperature at this large scale (SeedClim grid) -> add a plot with the correlation of soil and air temp across the grid in appendix
Data
SeedClim Data 2009 - 2018, maybe 2019 (2010 is missing)
Use FunCaB data from 2016 and 2018, but only few subplots are analysed in these years. Check if results change without these years.
RTC control plots from all years?
MO plots 2009 – 2010 (ask Lise)
SeedClim seed sowing control plots (2010 or 2011) -> ask Eric or Joachim

Check if there are there species that never flower at all, or very rare. Which species are those? Species have to occur in at least 3 years. Remove plants at the site level that never flowered. Remove shrubs, because they are not common. Only including graminoids and forbs.
Need some species level analysis. Compare species with different strategies:
For now focus on including biogeography in the model/analysis.
Alpine vs. generalist -> biogeography

Another approach would be to look at strategies, species level/groups. I.e. look at 3 groups: biogeography, reproductive strategy and growth strategy and how this affects fertility. Do a multivariate analysis separate for each of these groups (e.g. species vs. reproductive traits) and see if we get 1-2 axis that are important and explain the strategies.
Annual vs. perennial
Early vs. late flowering
Clonality within the perennials. Clonality: they do not need to flower every year or they have the resources to flower every year
Seed bank
Pollination strategy: wind vs. insect -> might reflect grass vs. forb (only Thalictrum is wind?)
growth traits: clonal traits, leaf traits



## REFERENCES
